valr: Reproducible genome interval analysis in R

### 2017-06-15 inst/paper/main.tex
version to send to all authors

% 300 words
\begin{abstract}
New tools for genome interval analysis enable analysis of large genomic data sets. The size and complexity of genomic data continues to increase highlighting the need for novels tools for exploratory data analysis. \texttt{valr} enables flexible and efficient genomic interval analysis in the RStudio environment. \texttt{valr} can be used for interactive exploratory analyses or incorporated into existing analysis pipelines.

@@

A routine task in bioinformatics is determining the relationships between sets of genomic intervals. Frequent tasks include for example, identifying SNPs located within geneic regions, annotating regions enriched for specific nucleic acid binding proteins, or computing read density across a set of exons. Command-line tools for interval analysis such as BEDtools \cite{quinlan_bedtools:_2010} and BEDOPS \cite{neph_bedops:_2012} enable analyses of genome-wide datasets and are central components of analysis pipelines. A common workflow for genomic analysis begins with preprocessing interval analyses via command-line tools, followed by importing data into R for visualization and further statistical analyses. However, this workflow separation hinders exploratory data analysis, and the development reproducible research workflows built in the RMarkdown framework.

Previous R packages developed for interval analysis include IRanges \cite{lawrence_software_2013}, bedr \cite{haider_bedr_2016}, and GenometriCorr \cite{favorov_exploring_2012}. IRanges is a Bioconductor package that provides specialized interval classes and methods to perform interval arithmetic in R and is used by many Bioconductor packages. In contrast, bedr is a CRAN distributed package that provides wrapper R functions to call BEDTools, BedOPS, and tabix command-line utilities providing out-of-memory support for interval analysis in R. GenometriCorr provides a set of statistical tests to determine the relationships between interval sets using IRanges data structures. These packages provide important functionality for processing and statistical inference of interval data, but they do not easily integrate with interactive data visualization features developed for R, such as Shiny\cite{chang_shiny_2017}, and the RStudio IDE. Additionally, the R S4 data structures implemented by IRanges do not easily interface with the recent advances provided by the popular tidyverse suite of data processing and visualization tools (dplyr, purrr, broom, and ggplot2) \cite{wickham_tidyverse_2017}. We therefore sought to develop an flexible R package for genomic interval arithmetic built to incorporate new R programming, visualization, and interactivity features.

@@

\subsection*{Implementation}
valr is an R package that makes extensive use of dplyr, a flexible and high-performance framework for data manipulation in R \cite{wickham_dplyr_2016}. Compute intensive functions in valr are written in C++ using Rcpp to enable fluid interactive analysis of large datasets\cite{eddelbuettel_rcpp_2011}. Interval intersections and related operations utilize an interval tree algorithm to efficently process large datasets\cite{cormen_2001}. BED files are imported and handled in R as data\_frame objects, requiring minimal pre or post-processing to integrate with additional R packages or command line tools.

\subsection*{Operation}
valr is distributed as part of the CRAN R package repository and is compatible with Mac OS X, Windows, and major Linux operating systems. Package dependencies and system requirements are documented on the valr CRAN repository (\url{https://cran.r-project.org/web/packages/valr/index.html}).

\section*{Use Cases}
To demonstrate the functionality and utility of \textbf{valr}, presented below is a basic tutorial for using valr and additional common use cases for genomic interval analysis.

@@

\texttt{valr} provides a set of functions to read BED, BEDgraph, and VCF formats into R as convenient \texttt{tbl\_dfs()} dataframe objects. All tbls will have \texttt{chrom}, \texttt{start}, and \texttt{end} columns, and tbls from multi-column formats will have additional pre-determined column names. Standards methods for importing data (e.g. \texttt{read.table}, \texttt{readr::read\_tsv}) are also supported provided the contructed dataframes contain the requiste column names (\texttt{chrom}, \texttt{start}, \texttt{end}). Additionally \texttt{valr} supports connections to remote databases to programatically access the UCSC and Ensembl databases via the \texttt{db\_ucsc} and \texttt{db\_ensembl} functions.

@@

\subsubsection*{Simple example}

The functions in \texttt{valr} have similar names to their
\texttt{BEDtools} counterparts, and so will be familiar to users of
the \texttt{BEDtools} suite. Also, similar to pybedtools\cite{dale_pybedtools:_2011}, a python wrapper for \texttt{BEDTools}, \texttt{valr} has a terse syntax. For example, shown below is an example of how to find all intergenic SNPs within 1 kilobase of genes using \texttt{valr}.

@@

\subsubsection*{Visual documentation}\label{visual-documentation}
By conducting interval arthimetic entirely in R, \texttt{valr} is also an effective teaching tool for introducing interval analysis to early-stage analysts without requiring familiarity with both command-line tools and R. To aid in demonstrating the interval operations available in \texttt{valr}, we developed the \texttt{bed\_glyph()} tool which produces plots demonstrating the input and output of operations in \texttt{valr} in a manner similar to those found in the \texttt{BEDtools} documentation.
Shown below is the code required to produce glyphs displaying the results of intersecting \texttt{x} and \texttt{y} intervals with \texttt{bed\_intersect()} , and the result of merging \texttt{x} intervals with \texttt{bed\_merge()} (Figure 1 \ref{fig:Figure 1}).

@@

This demonstration illustrates how to use \texttt{valr} tools to perform
a ``meta-analysis'' of signals relative to genomic features. Here we to
analyze the distribution of histone marks surrounding transcription
start sites, using H3K4Me3 Chip-Seq data downloaded from ENCODE (Link to data here?).

First we load libraries and relevant data.

@@

Then we generate 1 bp intervals to represent transcription start sites
(TSSs). We focus on \texttt{+} strand genes, but \texttt{-} genes are
easily accomodated by filtering them and using
\texttt{bed\_makewindows()} with \texttt{reverse}d window numbers.

@@

Here we examine the extent of overlap of repeat classes with exons in
the human genome (on \texttt{chr22} only, for simplicity) using the
jaccard similarity index. \texttt{bed\_jaccard()} implements the jaccard
test to examine the similarity between two sets of genomic intervals.
Using \texttt{bed\_shuffle()} and \texttt{replicate()} we generate a
\texttt{data\_frame} containing 100 sets of randomly placed intervals
then calculate the jaccard index for each set to generate a
null-distribution of jaccard scores. Finally an empirical p-value is
then calculated from the null-distribution.

@@

\subsection*{Benchmarking against bedtools}
In order to ensure that valr performs fast enough to enable interactive analysis, key functionality is implemented in C++. To test the speed of major valr functions we generated two dataframes containing 1 million random 1 kilobase intervals derived from the human genome. Most of the major valr functions complete execution in less than 1 second, demonstrating that valr can process large interval datasets efficiently (Figure 3A\ref{fig:Figure 3}).

@@

\subsection{Reproducible Reports and Interactive Visualizations}
Command-line tools like \texttt{BEDtools} and \texttt{bedops} can be incorporated into reproducible workflows (e.g., with
\href{https://bitbucket.org/snakemake/snakemake/wiki/Home}{\texttt{snakemake}}\cite{koster_snakemakescalable_2012}),
but it is cumbersome to transistion from command-line tools to exploratory analysis and plotting software. RMarkdown documents are plain text files, amenable to version control, which provide an interface to generate feature rich PDF and HTML reports that combine text, executable code, and figures in a single document. \texttt{valr} can be used in RMarkdown documents to provide rapid documentation of exploratory data analyses and generate reproducible work-flows for data processing.  Moreover, new features in RStudio, such as notebook viewing, and multiple language support enable similar functionality to another popular notebook platform \texttt{jupyter notebooks}.

Additionally, \texttt{valr} seamlessly integrates into R shiny applications allowing for complex interactive visualizations relating to genomic interval analyses. We have developed a shiny application available on github \href{https://github.com/rnabioco/valrdata} that explores CHiP-Seq signal density surrounding transcription start sites and demonstrates the ease of implementing \texttt{valr} to power dynamic visualizations, (e.g ).

\section*{Summary} % Optional - only if NO new datasets are included
valr provides a flexible framework for interval arithmetic in R/Rstudio. valr functions are written with a simple and terse syntax that promotes flexible interactive analysis. Additionally by providing an easy-to-use interface for interval arithmetic in R, valr is also a useful teaching tool to introduce the analyses necessary to investigate correlations between genomic intervals, without requiring familiarity with the command-line. We envison that \texttt{valr} will help researchers quickly and reproducibly analyze genome interval datasets.

\section*{Software availability}
\begin{enumerate}
\item URL valr can be installed via CRAN using "install.packages('valr')".
\item valr is maintained at http://github.com/rnabioco/valr
\item valr source code is available at http://github.com/rnabioco/valr
\item Latest stable version of source code is at: https://github.com/rnabioco/valr/archive/v0.2.0.tar.gz TO UPDATE
\item Software license: MIT
\end{enumerate}

@@

JR conceptualized the package and provided direction for software development. JR, KR, AG, and RS developed and tested the package.  YY and CB contributed workflows and benchmarks for the package. All authors wrote the manuscript.

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=0.5\textwidth]{figure2.pdf}
\caption{\label{fig:Figure 2}Meta-analysis of signals relative to genomic features with valr. \textnormal{(A) Summarized coverage of human H3K4Me3 Chip-Seq coverage across positive strand transcription start sites. Data presented +/- SEM.}}
\end{figure}

%%%% Figure 3 %%%%
\begin{figure}[!htb]
\centering
\includegraphics[width=1\textwidth]{figure3.pdf}
\caption{\label{fig:Figure 3} Performance of major valr functionality. \textnormal{(A)  Timings calculated by performing 10 repetitions of indicated functions on dataframes containing 1 million random 1 kilobase x/y intervals generated using \texttt{bed\_random()}. (B) Timings for executing functions in BEDTools or equivalent functions in valr using the same interval sets as in (A). All BEDTools functions were written to /dev/null, and were timed using GNU \texttt{time}. Timings for valr functions in (B) include times for reading files using \texttt{read\_bed()} and were timed using the \texttt{microbenchmark} package.}}
% * <jay.hesselberth@gmail.com> 2017-06-07T13:17:50.374Z:
%
% No bar charts! Make this a boxplot ideally.
%
% ^.
\end{figure}

>>>>

% 300 words
\begin{abstract}
New tools for reproducible exploratory data analysis of large data sets are important to address the rising size and complexity of genomic data. We developed the \texttt{valr} R package to enable flexible and efficient genomic interval analysis. \texttt{valr} leverages new tools available in the "tidyverse" including \texttt{dplyr}. Benchmarks of \texttt{valr} show it performs similar to \texttt{BEDtools} and can be used for interactive  analyses and incorporated into existing analysis pipelines.

@@

	A routine bioinformatic task is the analysis of the relationships between sets of genomic intervals including the identification of DNA variants within genetic regions, annotation of regions enriched for nucleic acid binding proteins, and computation of read density within a set of exons. Command-line tools for interval analysis such as BEDtools \cite{quinlan_bedtools:_2010} and BEDOPS \cite{neph_bedops:_2012} enable analyses of genome-wide datasets and are keys components of analysis pipelines. Analyses with these tools commonly combine processing intervals on the command-line with visualization and statistical analysis in R. However, the need to master both the command-line and R hinders exploratory data analysis, and the development reproducible research workflows built in the RMarkdown framework.

	Existing R packages developed for interval analysis include IRanges \cite{lawrence_software_2013}, bedr \cite{haider_bedr_2016}, and GenometriCorr \cite{favorov_exploring_2012}. IRanges is a Bioconductor package that provides  interval classes and methods to perform interval arithmetic and is used by many Bioconductor packages. bedr is a CRAN-distributed package that provides wrapper R functions to call the BEDTools, BedOPS, and tabix command-line utilities, providing out-of-memory support for interval analysis. Finally,  GenometriCorr provides a set of statistical tests to determine the relationships between interval sets using IRanges data structures. These packages provide  functionality for processing and statistical inference of interval data, but they do not easily integrate with interactive data visualization features developed for R including Shiny\cite{chang_shiny_2017}, and the RStudio IDE. Moreover, the S4 data structures implemented by IRanges do not easily interface with the recent advances provided by the popular tidyverse suite of data processing and visualization tools (e.g., dplyr, purrr, broom, and ggplot2) \cite{wickham_tidyverse_2017}. We sought to develop an flexible R package for genomic interval arithmetic built to incorporate new R programming, visualization, and interactivity features.

@@

\subsection*{Implementation}
valr is an R package that makes extensive use of dplyr, a flexible and high-performance framework for data manipulation in R \cite{wickham_dplyr_2016}. Compute intensive functions in valr are written in C++ using Rcpp to enable fluid interactive analysis of large datasets\cite{eddelbuettel_rcpp_2011}. Interval intersections and related operations use an interval tree algorithm to efficiently search for overlapping intervals\cite{cormen_2001}. BED files are imported and handled in R as data\_frame objects, requiring minimal pre or post-processing to integrate with additional R packages or command line tools.

\subsection*{Operation}
valr is distributed as part of the CRAN R package repository and is compatible with Mac OS X, Windows, and major Linux operating systems. Package dependencies and system requirements are documented in the valr CRAN repository (\url{https://cran.r-project.org/web/packages/valr/index.html}).

\section*{Use Cases}
To demonstrate the functionality and utility of \texttt{valr}, presented below is a basic tutorial for using valr and additional common use cases for genomic interval analysis.

@@

\texttt{valr} provides a set of functions to read BED, BEDgraph, and VCF formats into R as convenient \texttt{tibble} data\_frame objects. All tbls have \texttt{chrom}, \texttt{start}, and \texttt{end} columns, and tbls from multi-column formats have additional pre-determined column names. Standards methods for importing data (e.g. \texttt{read.table}, \texttt{readr::read\_tsv}) are also supported provided the constructed dataframes contain the requisite column names (\texttt{chrom}, \texttt{start}, \texttt{end}). Additionally \texttt{valr} supports connections to remote databases to programatically access the UCSC and Ensembl databases via the \texttt{db\_ucsc} and \texttt{db\_ensembl} functions.

@@

\subsubsection*{Example of combining valr tools}

The functions in \texttt{valr} have similar names to their \texttt{BEDtools} counterparts, and so will be familiar to users of the \texttt{BEDtools} suite. Also, similar to pybedtools\cite{dale_pybedtools:_2011}, a python wrapper for \texttt{BEDTools}, \texttt{valr} has a terse syntax. For example, shown below is a demonstration of how to find all intergenic SNPs within 1 kilobase of genes using \texttt{valr}.

@@

\subsubsection*{Visual documentation}\label{visual-documentation}
By conducting interval arithmetic entirely in R, \texttt{valr} is also an effective teaching tool for introducing interval analysis to early-stage analysts without requiring familiarity with both command-line tools and R. To aid in demonstrating the interval operations available in \texttt{valr}, we developed the \texttt{bed\_glyph()} tool which produces plots demonstrating the input and output of operations in \texttt{valr} in a manner similar to those found in the \texttt{BEDtools} documentation.
Shown below is the code required to produce glyphs displaying the results of intersecting \texttt{x} and \texttt{y} intervals with \texttt{bed\_intersect()} , and the result of merging \texttt{x} intervals with \texttt{bed\_merge()} (Figure 1 \ref{fig:Figure 1}).

@@

This demonstration illustrates how to use \texttt{valr} tools to perform
a ``meta-analysis'' of signals relative to genomic features. Here we
analyze the distribution of histone marks surrounding transcription
start sites, using H3K4Me3 Chip-Seq data downloaded from ENCODE.

First we load packages and relevant data.

@@

Then we generate 1 bp intervals to represent transcription start sites
(TSSs). We focus on \texttt{+} strand genes, but \texttt{-} genes are
easily accommodated by filtering them and using
\texttt{bed\_makewindows()} with \texttt{reversed} window numbers.

@@

Here we examine the extent of overlap of repeat classes with exons in
the human genome (on \texttt{chr22} only, for simplicity) using the
jaccard similarity index. \texttt{bed\_jaccard()} implements the jaccard
test to examine the similarity between two sets of genomic intervals.
Using \texttt{bed\_shuffle()} and \texttt{replicate()} we generate a
\texttt{data\_frame} containing 100 sets of randomly selected intervals
then calculate the jaccard index for each set against the repeat intervals
to generate a null-distribution of jaccard scores. Finally, an empirical p-value is
calculated from the null-distribution.

@@

\subsection*{Benchmarking against bedtools}
In order to ensure that valr performs fast enough to enable interactive analysis, key functionality is implemented in C++. To test the speed of major valr functions we generated two data\_frames containing 1 million random 1 kilobase intervals derived from the human genome. Most of the major valr functions complete execution in less than 1 second, demonstrating that valr can process large interval datasets efficiently (Figure 3A\ref{fig:Figure 3}).

@@

\subsection{Reproducible Reports and Interactive Visualizations}
Command-line tools like \texttt{BEDtools} and \texttt{bedops} can be incorporated into reproducible workflows (e.g., with
\href{https://bitbucket.org/snakemake/snakemake/wiki/Home}{\texttt{snakemake}}\cite{koster_snakemakescalable_2012}),
but it is cumbersome to transition from command-line tools to exploratory analysis and plotting software. RMarkdown documents are plain text files, amenable to version control, which provide an interface to generate feature rich PDF and HTML reports that combine text, executable code, and figures in a single document. \texttt{valr} can be used in RMarkdown documents to provide rapid documentation of exploratory data analyses and generate reproducible work-flows for data processing.  Moreover, new features in RStudio, such as notebook viewing, and multiple language support enable similar functionality to another popular notebook platform \texttt{jupyter notebooks}.

Additionally, \texttt{valr} seamlessly integrates into R shiny applications allowing for complex interactive visualizations relating to genomic interval analyses. We have developed a shiny application available on github \href{https://github.com/rnabioco/valrdata} that explores ChiP-Seq signal density surrounding transcription start sites and demonstrates the ease of implementing \texttt{valr} to power dynamic visualizations.

\section*{Summary} % Optional - only if NO new datasets are included
valr provides a flexible framework for interval arithmetic in R/Rstudio. valr functions are written with a simple and terse syntax that promotes flexible interactive analysis. Additionally by providing an easy-to-use interface for interval arithmetic in R, valr is also a useful teaching tool to introduce the analyses necessary to investigate correlations between genomic intervals, without requiring familiarity with the command-line. We envision that \texttt{valr} will help researchers quickly and reproducibly analyze genome interval datasets.

\section*{Software availability}
\begin{enumerate}
\item URL valr can be installed via CRAN using "install.packages('valr')".
\item valr is maintained at http://github.com/rnabioco/valr
\item valr source code is available at http://github.com/rnabioco/valr
\item Latest stable version of source code is at: https://github.com/rnabioco/valr/archive/v0.3.0.tar.gz
\item Software license: MIT
\end{enumerate}

@@

JR conceptualized the package and directed software development. JR, KR, AG, and RS developed and tested the package.  YY and CB contributed workflows and benchmarks for the package. All authors wrote the manuscript.

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=0.5\textwidth]{figure2.pdf}
\caption{\label{fig:Figure 2}Meta-analysis of signals relative to genomic features with valr. \textnormal{(A) Summarized coverage of human H3K4Me3 Chip-Seq coverage across positive strand transcription start sites on chromosome 22. Data presented +/- SEM.}}
\end{figure}

%%%% Figure 3 %%%%
\begin{figure}[!htb]
\centering
\includegraphics[width=1\textwidth]{figure3.pdf}
\caption{\label{fig:Figure 3} Performance of major valr functionality. \textnormal{(A)  Timings calculated by performing 10 repetitions of indicated functions on dataframes preloaded in R containing 1 million random 1 kilobase x/y intervals generated using \texttt{bed\_random()}. (B) Timings for executing functions in BEDTools or equivalent functions in valr using the same interval sets as in (A) written to files. All BEDTools function outputs were written to /dev/null/, and were timed using GNU \texttt{time}. Timings for valr functions in (B) include times for reading files using \texttt{read\_bed()} functions and were timed using the \texttt{microbenchmark} package.}}
\end{figure}
